Effectiveness of COVID-19 Booster doses in England: a sequential trials study in OpenSAFELY

### 2022-02-15 manuscript/draft-manuscript.Rmd

---
title: "Effectiveness of COVID-19 Booster doses in England: a sequential trials study in OpenSAFELY"
output:
  bookdown::html_document2:
    number_sections: no
    toc: no
  html_document:
    keep_md: yes
    self_contained: yes
  word_document: default
  pdf_document:
    toc: no
bibliography: references.bib
csl: vancouver.csl
link-citations: yes
zotero: yes
header-includes:
- \usepackage{float} 
- \floatplacement{figure}{H}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \captionsetup[table]{labelformat=empty}
---

@@

# where are the outputs (ie the inputs for this manuscript!) saved?
output_dir_os <- here("output", "manuscript-objects")
#output_dir_os <- here("released_output", "manuscript-objects")

@@

flowchart <- read_csv(fs::path(output_dir_os, "flowchart.csv"))
cohortsummary <- read_csv(fs::path(output_dir_os, "cohortsummary.csv"))
matchsummary <- read_csv(fs::path(output_dir_os, "matchsummary.csv"))
matchsummary_treated <- read_csv(fs::path(output_dir_os, "matchsummary_treated.csv")) %>% filter(treated == 1L)
matchsummary_control <- read_csv(fs::path(output_dir_os, "matchsummary_treated.csv")) %>% filter(treated == 0L)
matchcoverage <- read_csv(fs::path(output_dir_os, "matchcoverage.csv"))
incidences <- read_csv(fs::path(output_dir_os, "incidence.csv")) %>% mutate(events=events_1+events_0)

@@

hazards_vax12_type <- read_csv(fs::path(output_dir_os, "vax12_type", "effects.csv")) %>%
  mutate(
    subgroup_descr = fct_inorder(subgroup_descr),
    subgroup_level_descr = fct_inorder(subgroup_level_descr),
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    #model_name = fct_inorder(str_wrap(model_name, 15)),
    hr_display = paste0(number_format(accuracy=0.01)(hr), " (", number_format(accuracy=0.01)(hr.ll), "-", number_format(accuracy=0.01)(hr.ul), ")")
  )

@@

William J Hulme^1^, Elizabeth J Williamson^2^, Amelia Green^1^, Elsie Horne^4,5^, Linda Nab^1^, Ruth Keogh^2^, Edward Parker^2^, Jonathan AC Sterne^4,5^, Miguel Hernán^6,7^, Ben Goldacre^1^ + OpenSAFELY collaborative.

<!-- Krishnan Bhaskaran^2^, Helen I McDonald^2^, Christopher T Rentsch^2^, Anna Schultze^2^, John Tazare^2^, Helen J Curtis^1^, Alex J Walker^1^, Laurie Tomlinson^2^, Tom Palmer^4,5^,  Brian MacKenna^1^, Caroline E Morton^1^, Amir Mehrkar^1^, Louis Fisher^1^, Seb Bacon^1^, Dave Evans^1^, Peter Inglesby^1^, George Hickman^1^, Simon Davy^1^, Tom Ward ^1^, Richard Croker^1^, Rosalind M Eggo^2^, Angel YS Wong^2^, Rohini Mathur^2^, Kevin Wing^2^, Harriet Forbes^2^, Daniel Grint^2^, Ian J Douglas^2^, Stephen JW Evans^2^, Liam Smeeth^2^, Chris Bates^3^, Jonathan Cockburn^3^, John Parry^3^, Frank Hester^3^, Sam Harper^3^,  -->

@@

3\. TPP, TPP House, 129 Low Lane, Horsforth, Leeds, LS18 5PX

4\. Population Health Sciences, University of Bristol, Oakfield House, Oakfield Grove, Bristol, BS8 2BN, UK

5\. NIHR Bristol, Biomedical Research Centre, Bristol, UK

@@

There were `r label_number(1, big.mark=",")(filter(matchsummary_treated, treatment=="pfizer") %>% pull(n))` people boosted with BNT162b2 and `r label_number(1, big.mark=",")(filter(matchsummary_treated, treatment=="moderna") %>% pull(n))` people boosted with mRNA-1273 who were successfully matched to an unboosted control in the study period.

@@

In this cohort of fully vaccinated people in England, a booster dose was estimated to be highly effective at reducing.... , though there was evidence of waning effects. Heterologous boosting ...

@@

*Objectives* To estimate the effectiveness of the BNT162b2 and mRNA-1273 mRNA-1273 booster COVID-19 vaccines against infection and COVID-19 disease.

*Design* Matched sequential trial design, emulating a series of parallel two-arm RCTs.

*Setting* Linked primary care, hospital, and COVID-19 surveillance records available within the OpenSAFELY-TPP research platform.

*Participants* `r label_number(1, big.mark=",")(filter(flowchart, crit=="c4") %>% pull(n))` people with a complete primary COVID-19 vaccine course, registered with a GP practice using the TPP SystmOne clinical information system in England.

*Interventions* A booster vaccination with either BNT162b2 or mRNA-1273 administered as part of the national COVID-19 booster vaccine roll-out.

*Main outcome measures* Recorded SARS-CoV-2 infection, COVID-19-related accident and emergency attendance, unplanned COVID-19-related hospital admission, and COVID-19-related death.

*Conclusions*

@@

In the vast majority of cases, BNT162b2 and mRNA-1273 vaccines were used, and administered without consideration of the type of primary course. BNT162b2 was used from 16 September 2022, and mRNA-1273 from X onwards. Consequently, available follow-up of BNT162b2 recipients is longer, and so in this early analysis we restrict to effectiveness of this vaccine. An analysis of mRNA-1273 booster effectiveness is planned when sufficient follow-up has accrued.

@@

We used a sequential trials approach designed to emulate a sequence of daily two-arm parallel randomised controlled trials. For each trial, the treatment arm comprised people who received a BNT162b2 booster dose on that day, and the control arm comprised the same number of randomly selected unboosted people, matched on key characteristics described below.

@@

## Outcomes

@@

<!-- COVID-19 A&E attendances were identified using HES emergency care records with U07.1 ("COVID-19, virus identified") or U07.2 ("COVID-19, virus not identified") ICD-10 diagnosis codes @emergenc. -->

COVID-19 related A&E attendance and critical/intensive care admission we also considered but there were doubts over data the reliability of event ascertainment in these cases.

## Additional variables

Participant characteristics used describe trial participants and for confounder adjustment include: age, sex (male or female), English Index of Multiple Deprivation (IMD, grouped by quintiles), ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census), NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West), number of conditions in the clinically "at risk" classification, as per national prioritisation guidelines, the number of SARS-CoV-2 tests in the 6 months prior to the study start date (via SGSS), rurality (urban conurbation, urban city or town, rural town or village), evidence of prior SARS-CoV-2 infection (positive SARS-CoV-2 test, "probable" infection documented in primary care, or COVID-19 hospitalisation), learning disabilities, severe mental illness, undergoing an unplanned hospital admission, undergoing a planned hospital admission. All characteristics were ascertained as at the trial recruitment date (or study start date if assumed to be fixed from that date onwards).

@@

Time zero was the trial recruitment day, and trial participants were followed up for no more than 10 weeks from this day. Follow-up was censored earlier than this at `r format(as.Date(study_dates$studyend_date), "%d %B %Y")`, death, or GP practice de-registration. Additionally for each matched pair, follow-up for both the boosted and unboosted person was censored if the control is boosted.

@@

## Subgroup analyses

@@

Participant characteristics at trial recruitment between trial arms were often well-balanced (Table 1), with some notable exceptions. Unboosted controls were typically... These unbalanced characteristics were adjusted for in the Cox regression models.

@@

```{r table2, echo=FALSE, warning=FALSE, message=FALSE}
incidences %>%
  filter(
    fup_period == "All",
    outcome %in% c("postest", "covidadmitted", "coviddeath")
  ) %>%
  left_join(
    matchsummary %>% select(treatment, eligible_1=n_eligible, recruit_prop_1 = prop_matched_eligible),
    by=c("treatment")
  ) %>%
  select(
    treatment_descr, 
    outcome_descr,
    eligible_1,
    n_1,
    recruit_prop_1,
    q_1,
    rate_1,
    q_0, 
    rate_0,
    rr, 
    rrCI
  ) %>%
  gt(
    groupname_col = c("treatment_descr"),
  ) %>%
  cols_label(
    #treatment_descr = "Treatment",
    outcome_descr = "Outcome",

    eligible_1 = "Eligible",
    n_1 = "Recruited",
    recruit_prop_1 = "(%)",

@@

    rr = "Incidence rate ratio",
    rrCI = "95% CI"
  ) %>%
  tab_spanner(
    label = "Boosted",
    columns = ends_with("1")
  ) %>%
  tab_spanner(
    label = "Unboosted",
    columns = ends_with("0")
  ) %>%
  fmt_number(
    columns = c("rate_0", "rate_1", "rr"),
    decimals = 2
  ) %>%
  fmt_integer(
    columns = c("n_1", "eligible_1"),
    sep_mark = ","
  ) %>%
  fmt_number(
    columns = "recruit_prop_1",
    scale_by = 100,
    decimals = 1,

  ) %>%
  fmt_missing(
    everything(),
    missing_text="--"
  ) %>%
  cols_align(
    align = "right",
    columns = everything()
  )

@@

Trial participants contributed `r filter(incidences, treatment=="pfizer", outcome=="postest", fup_period=="all") %>% pull(events)` positive SARS-CoV-2 tests, `r filter(incidences, treatment=="pfizer", outcome=="covidadmitted", fup_period=="all") %>% pull(events)` COVID-19 hospital admissions, and `r filter(incidences, treatment=="pfizer", outcome=="coviddeath", fup_period=="all") %>% pull(events)` COVID-19 deaths (Table 2).

@@

During the fifth and sixth week after recruitment (days 28-41) the estimated hazard ratio comparing a BNT162b2 booster dose to no dose was `r filter(hazards_none, outcome=="postest", treatment=="pfizer", model==3, term_left==28) %>% pull(hr_display)` for a positive test, `r filter(hazards_none, outcome=="covidadmitted", treatment=="pfizer", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 hospital admission, and `r filter(hazards_none, outcome=="coviddeath", treatment=="pfizer", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 death (Figure 2).

@@

For the same period, the estimated effectiveness for BNT162b2 and ChAdOx1-S primary courses respectively were `r filter(hazards_vax12_type, outcome=="postest", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="postest", treatment=="pfizer", subgroup_level=="az-az", model==3, term_left==28) %>% pull(hr_display)` for a positive test, `r filter(hazards_vax12_type, outcome=="covidadmitted", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="covidadmitted", treatment=="pfizer", subgroup_level=="az-az", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 hospital admission, and `r filter(hazards_vax12_type, outcome=="coviddeath", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="coviddeath", treatment=="pfizer", subgroup_level=="az-az", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 death (Figure 2).

@@

  plot_hazards <-
    hazards3_vax12_type %>%
    filter(outcome %in% c("postest", "covidadmitted", "coviddeath")) %>%
    ggplot()+
    geom_point(aes(y=hr, x=term_midpoint, colour=model_descr), position = position_dodge(width = 1.8))+
    geom_linerange(aes(ymin=hr.ll, ymax=hr.ul, x=term_midpoint, colour=model_descr), position = position_dodge(width = 1.8))+
    geom_hline(aes(yintercept=1), colour='grey')+
    facet_grid(rows=vars(outcome_descr), cols=vars(subgroup_level_descr), switch="y")+
    scale_y_log10(
      breaks=y_breaks,
      limits = c(0.01, 1.5),
      oob = scales::oob_keep,
      sec.axis = sec_axis(
        ~(1-.),
        name="Effectiveness",
        breaks = 1-(y_breaks),
        labels = function(x){formatpercent100(x, 1)}
      )
    )+
    scale_x_continuous(breaks=postbaselinecuts, limits=c(min(postbaselinecuts), max(postbaselinecuts)+1), expand = c(0, 0))+
    scale_colour_brewer(type="qual", palette="Set2", guide=guide_legend(ncol=1))+
    labs(
      y="Hazard ratio",
      x="Days since booster dose",
      colour=NULL
     ) +
    theme_bw()+
    theme(
      panel.border = element_blank(),
      axis.line.y = element_line(colour = "black"),
      axis.line.y.right = element_blank(),

@@

This observational study of fully vaccinated adults in England estimated high protection of BNT162b2 booster vaccination against all studied outcomes. 
Protection appeared higher in those receiving a X-X primary course.
There were apparent waning effects after X weeks which is unlikely to explained by increased transmissibility of the Omicron variant given Omicron was not yet dominant in the UK by the end of the follow-up window (ref).

@@

We used routinely-collected health records with comprehensive coverage of primary care, hospital admissions, COVID-19 testing and vaccination, and death registrations to study over X million fully vaccinated English residents ...

This study considers effectiveness in a population where Alpha was the predominant circulating variant (ref gov.uk reports). Re-running the analysis against a backdrop of the now globally dominant Omicron variant will be undertaken where valid concurrent comparisons between boosted and unboosted people can be made.

@@

@@

[pfizer phase 3 booster trial, in pfizer first course], [ cov-boost trial]

A phase III trial assessing BNT162b2 booster efficacy in two-dose BNT162b2 recipients found.. (ref)

A number of studies have estimated COVID-19 vaccine booster effectiveness in observational data.

@@

This study found high protection against severe COVID-19 disease

<!-- \newpage -->

<!-- # Summary -->

<!-- ### What is already known -->

<!-- ### What this study adds -->

@@

## Guarantor

>>>>

---
title: "Effectiveness of COVID-19 Booster doses in England: a sequential trials study in OpenSAFELY"
output:
  word_document: default
  html_document:
    keep_md: yes
    self_contained: yes
  bookdown::html_document2:
    number_sections: no
    toc: no
  pdf_document:
    toc: no
bibliography: references.bib
csl: vancouver.csl
link-citations: yes
zotero: yes
header-includes:
- \usepackage{float} 
- \floatplacement{figure}{H}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \captionsetup[table]{labelformat=empty}
---

@@

# where are the outputs (ie the inputs for this manuscript!) saved?
#output_dir_os <- here("output", "manuscript-objects")
output_dir_os <- here("released_output", "manuscript-objects")

@@

flowchart <- read_csv(fs::path(output_dir_os, "flowchart.csv"))
cohortsummary <- read_csv(fs::path(output_dir_os, "cohortsummary.csv"))
matchsummary <- read_csv(fs::path(output_dir_os, "matchsummary.csv"))
matchsummary_treated <- read_csv(fs::path(output_dir_os, "matchsummary_treated.csv")) %>% filter(treated == 1L)
matchsummary_control <- read_csv(fs::path(output_dir_os, "matchsummary_treated.csv")) %>% filter(treated == 0L)
matchcoverage <- read_csv(fs::path(output_dir_os, "matchcoverage.csv"))
incidences <- read_csv(fs::path(output_dir_os, "none", "incidence.csv")) %>% mutate(events=events_1+events_0)

@@

hazards_vax12_type <- read_csv(fs::path(output_dir_os, "vax12_type", "effects.csv")) %>%
  mutate(
    subgroup_descr = fct_inorder(subgroup_descr),
    subgroup_level = fct_inorder(subgroup_level),
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    #model_name = fct_inorder(str_wrap(model_name, 15)),
    hr_display = paste0(number_format(accuracy=0.01)(hr), " (", number_format(accuracy=0.01)(hr.ll), "-", number_format(accuracy=0.01)(hr.ul), ")")
  )

@@

William J Hulme^1^, Elizabeth J Williamson^2^, Amelia Green^1^, Elsie Horne^3,4^, Helen Curtis^1^, Linda Nab^1^, Milan Wiedemann^1^, Ruth Keogh^2^, Edward Parker^2^, Venexia Walker^3,4^, Jonathan AC Sterne^3,4^, Miguel Hernán^6,7^, Ben Goldacre^1^ + OpenSAFELY collaborative.

<!-- Krishnan Bhaskaran^2^, Helen I McDonald^2^, Christopher T Rentsch^2^, Anna Schultze^2^, John Tazare^2^, Helen J Curtis^1^, Alex J Walker^1^, Laurie Tomlinson^2^, Tom Palmer^3,4^,  Brian MacKenna^1^, Caroline E Morton^1^, Amir Mehrkar^1^, Louis Fisher^1^, Seb Bacon^1^, Dave Evans^1^, Peter Inglesby^1^, George Hickman^1^, Simon Davy^1^, Tom Ward ^1^, Richard Croker^1^, Rosalind M Eggo^2^, Angel YS Wong^2^, Rohini Mathur^2^, Kevin Wing^2^, Harriet Forbes^2^, Daniel Grint^2^, Ian J Douglas^2^, Stephen JW Evans^2^, Liam Smeeth^2^, Chris Bates^5^, Jonathan Cockburn^5^, John Parry^5^, Frank Hester^5^, Sam Harper^5^,  -->

@@

3\. Population Health Sciences, University of Bristol, Oakfield House, Oakfield Grove, Bristol, BS8 2BN, UK

4\. NIHR Bristol, Biomedical Research Centre, Bristol, UK

5\. TPP, TPP House, 129 Low Lane, Horsforth, Leeds, LS18 5PX

@@

We matched a total of `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_matched))` BNT162b2 booster recipients to unboosted controls with only two doses. During the 28-41 days the estimated hazard ratio (95% CI) comparing a BNT162b2 booster dose to two doses only, was `r filter(hazards_none, outcome=="postest", treatment=="pfizer", model==3, term_left==28) %>% pull(hr_display)` for a positive test, `r filter(hazards_none, outcome=="covidadmitted", treatment=="pfizer", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 hospital admission, and `r filter(hazards_none, outcome=="coviddeath", treatment=="pfizer", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 death.

@@

BNT162b2 boosting was estimated to be highly effective in protecting against COVID-19 related hospitalisation and death.

@@

<!-- *Objectives* To estimate the effectiveness of the BNT162b2 and mRNA-1273 mRNA-1273 booster COVID-19 vaccines against infection and COVID-19 disease. -->

<!-- *Design* Matched sequential trial design, emulating a series of parallel two-arm RCTs. -->

<!-- *Setting* Linked primary care, hospital, and COVID-19 surveillance records available within the OpenSAFELY-TPP research platform. -->

<!-- *Participants* `r label_number(1, big.mark=",")(filter(flowchart, crit=="c4") %>% pull(n))` people with a complete primary COVID-19 vaccine course, registered with a GP practice using the TPP SystmOne clinical information system in England. -->

<!-- *Interventions* A booster vaccination with either BNT162b2 or mRNA-1273 administered as part of the national COVID-19 booster vaccine roll-out. -->

<!-- *Main outcome measures* Recorded SARS-CoV-2 infection, COVID-19-related accident and emergency attendance, unplanned COVID-19-related hospital admission, and COVID-19-related death. -->

<!-- *Results* -->

<!-- *Conclusions* -->

@@

In the vast majority of cases, BNT162b2 and mRNA-1273 vaccines were used, and administered without consideration of the type of primary course. BNT162b2 was used from 16 September 2021, and mRNA-1273 from 29 October 2021 onwards. Consequently, available follow-up of BNT162b2 recipients is longer, and so we restrict the analysis to the effectiveness of BNT162b2boosting only. An analysis of mRNA-1273 booster effectiveness is planned when sufficient follow-up has accrued.

@@

We used a sequential trials approach designed to emulate a sequence of daily two-arm parallel randomised controlled trials. For each trial, the treatment arm comprised people who received a BNT162b2 booster dose on that day, and the control arm comprised the same number of randomly selected unboosted people matched on key characteristics, as described below. This closely follows the approach taken in previous vaccine effectiveness studies in Israeli registry data @dagan2021 @barda2021

@@

## Exposures, outcomes, covariates

<!--# COVID-19 A&E attendances were identified using HES emergency care records with U07.1 ("COVID-19, virus identified") or U07.2 ("COVID-19, virus not identified") ICD-10 diagnosis codes @emergenc. -->

@@

<!--# COVID-19 related A&E attendance and critical/intensive care admission we also considered but there were doubts over data the reliability of event ascertainment in these cases. -->

Participant characteristics used describe trial participants and for confounder adjustment include: age, sex (male or female), English Index of Multiple Deprivation (IMD, grouped by quintiles), ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census), NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West), number of conditions in the clinically "at risk" classification, as per national prioritisation guidelines, the number of SARS-CoV-2 tests in the 6 months prior to the study start date (via SGSS), rurality (urban conurbation, urban city or town, rural town or village), evidence of prior SARS-CoV-2 infection (positive SARS-CoV-2 test, "probable" infection documented in primary care, or COVID-19 hospitalisation), learning disabilities, severe mental illness, undergoing an unplanned hospital admission, undergoing a planned hospital admission. All characteristics were ascertained as at the trial recruitment date.

@@

Time zero was the trial recruitment day, and trial participants were followed up for no more than 10 weeks from this day. Follow-up was censored earlier than this at `r format(as.Date(study_dates$studyend_date), "%d %B %Y")`, death, or GP practice de-registration. Additionally for each matched pair, follow-up for both the boosted and unboosted person was censored if the control received a booster dose.

@@

@@

Participant characteristics at trial recruitment between trial arms (Table 1) were typically well-balanced (necessarily so on matched characteristics), though with some exceptions, for instance unboosted controls had higher levels of deprivation, higher rates of learning disabilities and severe mental illness, and lower SARS-CoV-2 testing frequency prior to the study. These unbalanced characteristics were adjusted for in the Cox regression models.

@@

```{r table2, echo=FALSE, warning=FALSE, message=FALSE}
incidences %>%
  filter(
    fup_period == "All",
    outcome %in% c("postest", "covidadmitted", "coviddeath"),
    treatment == "pfizer"
  ) %>%
  left_join(
    matchsummary %>% select(treatment, n_eligible, prop_matched_eligible),
    by=c("treatment")
  ) %>%
  mutate(
    treatment_decr = paste0(
      treatment_descr, " - ",
      label_number(1, big.mark=",")(n_eligible), "eligible, ",
      label_number(1, big.mark=",")(n_1), "matched (",
      label_percent(0.1)(prop_matched_eligible), ")."
    )
  ) %>%
  select(
    treatment_descr, 
    outcome_descr,
    q_1,
    rate_1,
    q_0, 
    rate_0,
    rr, 
    rrCI
  ) %>%
  gt(
    groupname_col = c("treatment_descr"),
  ) %>%
  cols_label(
    #treatment_descr = "Treatment",
    outcome_descr = "Outcome",

@@

    rr = "Incidence rate ratio",
    rrCI = "95% CI"
  ) %>%
  tab_spanner(
    label = "Boosted",
    columns = ends_with("1")
  ) %>%
  tab_spanner(
    label = "Unboosted",
    columns = ends_with("0")
  ) %>%
  fmt_number(
    columns = c("rate_0", "rate_1", "rr"),
    decimals = 3
  ) %>%
  fmt_missing(
    everything(),
    missing_text="--"
  ) %>%
  cols_align(
    align = "right",
    columns = everything()
  )

@@

Trial participants contributed `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="postest", fup_period=="All") %>% pull(events))` positive SARS-CoV-2 tests, `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="covidadmitted", fup_period=="All") %>% pull(events))` COVID-19 hospital admissions, and `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="coviddeath", fup_period=="All") %>% pull(events))` COVID-19 deaths (Table 2).

@@

During the 28-41 days the estimated hazard ratio (95% CI) comparing a BNT162b2 booster dose to two doses only, was `r filter(hazards_none, outcome=="postest", treatment=="pfizer", model==3, term_left==28) %>% pull(hr_display)` for a positive test, `r filter(hazards_none, outcome=="covidadmitted", treatment=="pfizer", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 hospital admission, and `r filter(hazards_none, outcome=="coviddeath", treatment=="pfizer", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 death (Figure 2).

@@

For the same period, the estimated hazard ratio (95% CI) for BNT162b2 and ChAdOx1-S primary courses respectively were `r filter(hazards_vax12_type, outcome=="postest", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="postest", treatment=="pfizer", subgroup_level=="az-az", model==3, term_left==28) %>% pull(hr_display)` for a positive test, `r filter(hazards_vax12_type, outcome=="covidadmitted", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="covidadmitted", treatment=="pfizer", subgroup_level=="az-az", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 hospital admission, and `r filter(hazards_vax12_type, outcome=="coviddeath", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3, term_left==28) %>% pull(hr_display)` and `r filter(hazards_vax12_type, outcome=="coviddeath", treatment=="pfizer", subgroup_level=="az-az", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 death (Figure 2).

@@

  plot_hazards <-
    hazards3_vax12_type %>%
    filter(
      outcome %in% outcomes,
      treatment %in% c("pfizer")
    ) %>%
    ggplot()+
    geom_point(aes(y=hr, x=term_midpoint, colour=model_descr), position = position_dodge(width = 1.8))+
    geom_linerange(aes(ymin=hr.ll, ymax=hr.ul, x=term_midpoint, colour=model_descr), position = position_dodge(width = 1.8))+
    geom_hline(aes(yintercept=1), colour='grey')+
    facet_grid(rows=vars(outcome_descr), cols=vars(subgroup_level), switch="y")+
    scale_y_log10(
      breaks=y_breaks,
      limits = c(0.01, 1.5),
      oob = scales::oob_keep,
      sec.axis = sec_axis(
        ~(1-.),
        name="Effectiveness",
        breaks = 1-(y_breaks),
        labels = function(x){formatpercent100(x, 1)}
      )
    )+
    scale_x_continuous(breaks=postbaselinecuts, limits=c(min(postbaselinecuts), max(postbaselinecuts)+1), expand = c(0, 0))+
    scale_colour_brewer(type="qual", palette="Set2", guide=guide_legend(ncol=1))+
    labs(
      y="Hazard ratio",
      x="Days since booster dose",
      colour=NULL
     ) +
    theme_bw()+
    theme(
      panel.border = element_blank(),
      axis.line.y = element_line(colour = "black"),
      axis.line.y.right = element_blank(),

@@

This observational study of over 8 million boosted adults in England and their matched controls estimated high protection of BNT162b2 booster vaccination against all studied outcomes using routinely-collected health records with comprehensive coverage of primary care, hospital admissions, COVID-19 testing and vaccination, and death registrations. Protection was similar between 2x BNT162b2 primary course recipients and 2x ChAdOx1-S recipients. There were apparent waning effects around 6 weeks after boosting, which is unlikely to explained by increased transmissibility of the Omicron variant given Omicron was not yet dominant in the UK by the end of the follow-up window @COVID19OmicronDaily, though changing testing behaviours over time between boosted and unboosted people may influence estimates.

@@

@@

This study considers effectiveness in a population where Delta was the predominant circulating variant (ref gov.uk reports). Re-running the analysis against a backdrop of the now globally dominant Omicron variant will be undertaken where valid concurrent comparisons between boosted and unboosted people can be made.

@@

<!-- [ cov-boost trial] -->

A phase III trial assessing BNT162b2 booster efficacy in two-dose BNT162b2 recipients without prior infection reported a relative efficacy of 95.6% (95% CI 89.3-98.6) against Delta infection @PfizerBioNTechAnnounce2021.

Recent studies in the UK have provided observational evidence for increased protection from the booster vaccine against symptomatic COVID-19 infection in comparison to second dose recipients using a test negative design. Andrews et al. reported estimates of relative vaccine effectiveness in the 14 days after a BNT162b2 booster dose in those aged over 50 years of 87.4% (95%CI 84.9-89.4) where the primary course was ChAdOx1-S and 84.4% (95%CI 82.8-85.8) for BNT162b2 @andrews2021. Sheikh et al.  reported comparable estimates of relative vaccine effectiveness for S gene positive symptomatic infection (83% (95%CI 81-84) 6-49 years, 88% (95%CI 86-89) ≥50 years) in the 14 days after a BNT162b2 or mRNA-1273 booster dose, but lower estimates for S gene negative symptomatic infection (56% (95%CI 51-60) 6-49 years, 57% (95%CI 52-62) ≥50 years).

<!-- @https://www.research.ed.ac.uk/en/publications/severity-of-omicron-variant-of-concern-and-vaccine-effectiveness- -->

<!-- Elsewhere, studies in Israel found .. -->

<!-- @1050236/technical-briefing-34-14-january-2022.pdf -->

@@

This study found high protection for BNT162b2 boosting against COVID-19 hospitalisation and death in the Delta era.

@@
